---
title: "Data Cleaning & Preparation for Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{data_cleaning_preparation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ManyEcoEvo)
```

## Data Cleaning
Analysts may report estimates on various scales, for example they may report values on the link or response scales, they may also ....
In order to proceed with effect size standardisation or other standardisation, some transformations or back-transformations need to be applied. 

- Anonymising submissions - `anonymise_teams()` / `anonymise_id_col()`
- back-transformation:
	- We asked for estimates to be provided on the `______` scale.
		- `clean_response_transformation()`
		- `assign_back_transformation_type()` 
			- Depends on what scale the values are reported on (link / response), what transformation has been applied to the response variable, and to the predictor variable. 

## Data Preprocessing

what is the distinction between processing and cleaning?

### Back-transforming response variables

in meta-analysis units need to be standardized because analysts use different response / outcome variables, and we may need to ensure that we are talking about units on the correct scale before we can standardize. 

| function name     | description |
| ----------------- | --------- |
| `log_back()`      |           |
| `logit_back()`    |           |
| `probit_back()`   |           |
| `inverse_back()`  |           |
| `square_back()`   |           |
| `cube_back()`     |           |
| `identity_back()` |           |
| `power_back()`    |           |
| `divide_back()`   |           |
| `square_root_back()`                  |           |



### Standardising effect sizes
- Standardisation of effect-sizes (fishers' Z), however other transformations could be applied using other packages if need be (Gurrindgi green meta-analsis handbook).
	- Predictions
		-  `pred_to_Z()` (data frame level), `Z_VZ_preds()`
	- Coefficients
		- `est_to_Zr()`

### Calculating Sorensen similarity index

- `apply_sorensen_calc()`
- `calculate_sorensen_diversity_index()` (also needs to be renamed)

### Box-cox transforming deviation from meta-analytic mean

### Excluding Data

- exclude_extreme_VZ()




