---
tags:
  - R
  - meta-analysis
  - metaresearch
  - ecology
  - evolutionary biology
  - many-analysts
authors:
  - name: Elliot Gould
    orcid: 0000-0003-0872-7098
    email: "elliot.gould@unimelb.edu.au"
    affiliations:
      - ref: aff1
  - name: Hannah S. Fraser
    orcid: 0000-0000-0000-0000
    affiliations:
      - ref: aff1
  - name: Shinichi Nakagawa
    orcid: 0000-0000-0000-0000
    affiliations:
      - ref: aff2
  - name: Timothy H. Parker
    orcid: 0000-0000-0000-0000
    affiliations:
      - ref: aff3
citation_author: Gould and Fraser et al.
affiliations:
  - id: aff1
    name: The University of Melbourne
  - id: aff2
    name: The University of New South Wales
  - id: aff3
    name: Whitman College
format:
  html:
    html-math-method: mathjax
title: "ManyEcoEvoR"
subtitle: "An R package for Analysing Many-analyst Style Data"
vignette: >
  %\VignetteIndexEntry{Vignette's Title}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
code-overflow: wrap
include: true
echo: true
message: false
warning: false
error: false
cache: true
bibliography: "vignettes.bib"
---

```{r}
#| echo: false
#| message: false
#| warning: false
#| cache: false
#| label: setup
library(ManyEcoEvo)
library(tidyverse)
library(metafor)
set.seed(1)
knitr::opts_chunk$set(
  # collapse = TRUE,
  comment = "#>",
  # fig.width = 7,
  # fig.height = 5,
  message = FALSE,
  warning = FALSE,
  include = TRUE,
  cache = TRUE,
  echo = TRUE,
  tidy = "styler"
)

```

## Summary

Many-analyst studies are a format of crowd-sourced scientific analysis [@Schweinsberg2021], where analysts simultaneously, but independently, research the same research question, using the same dataset [@Voracek2019; @Silberzahn2018]. Many-analyst studies demonstrate how subjective, often arbitrary and usually defensible analysis decisions may influence research results [@Silberzahn2018] due to either disagreement among researchers or due to analytic uncertainty about choices such as which observations to include and what data-analysis strategy is most appropriate [@Voracek2019]. Managing the analysis pipelines to implement many-analyst studies is complex and best implemented with dedicated software.

Typical scientific approaches proceed with a single team publishing only one or several ‘analysis specifications,’ or unique sequences of analysis choices. Yet, researchers routinely explore and implement multiple plausible analysis specifications, often without disclosing the full set of analytic options they explored, and so the effect of alternative defensible analytic choices is unknown. The growing body of many-analyst, and ‘multiverse’ or ‘specification analysis’ studies, in which a large decision space containing the full set of combinatorial alternatives is implemented and statistically analysed, reveals that there are a wide range of justifiable analytic options for any given study and the results from these analytic options can substantially alter the conclusions [@Botvinik-Nezer2020; @Simonsohn2020; @Olsson-Collentine2025].

*ManyEcoEvo* is an R package [@base] designed to *facilitate the analysis of* an existing many-analyst study in ecology and evolutionary biology, the ManyEcoEvo study [@Gould2025]. The purpose of *ManyEcoEvo* is to reproducibly implement the analysis and presentation of results for @Gould2025 such that the 246 analysts as well as other interested researchers may independently reproduce the study's findings. The package also aims to allow re-analyses and replications of @Gould2025, as well as facilitating the analysis of similar many-analyst style datasets.

The ManyEcoEvo study is a many-analyst study that aimed to empirically explore the variation in effect sizes and model predictions generated by analytical decisions of different researchers within ecology and evolutionary biology [@Gould2025]. 174 analyst teams, comprising 246 analysts, independently investigated the answers to two prespecified research questions using two previously unpublished datasets, one from conservation ecology, and one from evolutionary biology. Analysts investigating the evolutionary biology dataset examined the effect of sibling number and nestling growth of blue tits (*Cyanistes caeruleus*), while analysts investigating the conservation ecology dataset examined the relationship between grass cover and *Eucalyptus* tree seedling recruitment. The ManyEcoEvo study quantified the variation in results among analyses using meta-analysis and additionally sought to understand how features of studies such as, variable selection, random effects structures, qualitative and numeric ratings of the methods by peer-reviewers, influenced deviation from the meta-analytic mean.

All raw data sets and metadata provided to independent teams for analysis are publicly archived at [osf.io/qjzby](https://osf.io/qjzby) and [osf.io/hdv8m](https://osf.io/hdv8m), respectively. Analyst survey responses and results are provided in the package GitHub repository ([github.com/egouldo/ManyEcoEvo](https://github.com/egouldo/ManyEcoEvo)), which is also archived on Zenodo ([doi.org/10.5281/zenodo.10046153](https://doi.org/10.5281/zenodo.10046153)). *ManyEcoEvo* contains tools for data-cleaning and processing tasks encountered when working with many-analyst style data, functions for implementing and reproducing the analysis described in @Gould2025, establishes a structured workflow that facilitates re-analyses and replications of the study, as well as implementation for new many-analyst studies, including a planned extension of the ManyEcoEvo study.

## Statement of Need

The last ten years has seen the growing emergence of many-analyst projects, [e.g. @Silberzahn2018; @Botvinik-Nezer2020; @Huntington-Klein2021; @Schweinsberg2021; @breznau2022; @Hoogeveen2022; @coretta2023], predominantly in the disciplines of psychology and social and behavioural sciences. Given the increasing attention on the reproducibility crisis, it is expected that more of these types of studies will be published in the near future across other disciplines. Moreover, collaborative and distributed big team open science projects are increasingly common in ecology and evolutionary biology (e.g. NutNet [dragnetglobal.org](https://dragnetglobal.org) and DragNet [nutnet.org/index.php/](https://nutnet.org/index.php/){.uri}; ManyBirds [themanybirds.com](http://themanybirds.com){.uri}) as well as other disciplines [ManyBabies, @ManyBabies; ManyPrimates, @ManyPrimates] which pose computationally analogous data-manipulation and analysis problems to the many-analyst workflow addressed by the *ManyEcoEvo* approach.

The *ManyEcoEvo* package was created to implement the analysis for the 'Same Data, Many Analysts in Ecology and Evolutionary Biology' project [@Gould2025]. We opted to develop this research compendium as an R package as we find it best encapsulates the principles of 'clean code' for effective communication of the science [@Filazzola2022], and therefore facilitates future reproduction, replication and potential synthesis [@Marwick2018; @Vuorre2021].

The results of some many-analyst studies have been controversial, yielding re-analyses criticising the original implementation of some many-analyst studies. For example, @Auspurg2021 re-analysed part of the seminal 'Red-Card' many-analyst study examining the influence of whether referees were more likely to give red cards to soccer players with dark skin tone than those with light skin tone [@Silberzahn2018]. By providing the code and data for @Gould2025 in the *ManyEcoEvo* package, we allow similar re-analyses of the ‘Same Data, Many Analysts in Ecology and Evolutionary Biology’, wherein ManyEcoEvo can be used to re-run the analysis in @Gould2025 with slightly different methods, or subsets of data in order to verify and examine the robustness of our findings. We also provide infrastructure to support others performing many-analyst studies.

We are not aware of any R packages or software in other languages exclusively focused on facilitating many-analyst style studies. Existing many-analyst code if shared, is written as R scripts tailored to that study's analysis (e.g. [github.com/manybabies/mb1-analysis-public](https://github.com/manybabies/mb1-analysis-public), [github.com/many-speech-analyses/many_analyses](https://github.com/many-speech-analyses/many_analyses)). However, there are a range of existing packages with select functionality that addresses various components of many-analyst style analyses, namely meta-analysis, its associated preparation and visualisation, as well as the visualisation component of specification curve analysis / multiverse analysis. Our package unites the functions of these packages to provide a selection of tools that are customised to the needs of a many-analyst project. In a many-analyst study, multiple researchers submit their analyses and results. The submitted data must be cleaned and prepared for meta-analysis and other model analysis, then subjected to model checking, statistic extraction and visualisation [@Aczel2021].

We provide a suite of functions for data cleaning and processing tasks that we expect to be common to the sorts of data collected by many-analyst style studies. This was the most time and effort-intensive aspect of the analysis in @Gould2025, and we hope these functions will ease the burden on future researchers conducting many-analyst studies.

*ManyEcoEvo* provides custom functions to calculate standardised effect sizes in the form of Fishers' Z correlation scores in preparation for meta-analysis. While some existing functions exist within other packages [e.g. `DescTools::FishersZ()`, @DescTools; `DiagTest3Grp::FisherZ.var()`, @luo2012diagtest3grp], the method for calculating variance of the standardised effect size in @Gould2025 differed slightly, and we wished to avoid reliance on too many external packages [@wickham2023r, Chapter 10.1]. The underlying modularity of *ManyEcoEvo* means that users can supply their own custom functions or or functions provided by other software should they wish to employ different effect-size standardisation methods [see @harrer2021, see Chapter 3 R code to calculate various effect sizes].

The two most commonly used software packages for undertaking meta-analysis in R are *metafor* [@metafor] and *meta* [@Balduzzi2019; @lortie2020]. We provide functions for fitting the exact meta-analysis models used in @Gould2025, which were implemented with the *metafor* package [@metafor]. However, *ManyEcoEvo* can accept models fitted with alternative model-fitting functions for further visualisation and analysis.

After model-fitting and analysis, model checking and extraction of model fit and heterogeneity statistics is implemented using a range of existing packages, including: *performance* [@Ludecke2020], *broom.mixed* [@Bolker2022], *metafor* [@metafor]. *ManyEcoEvo* also draws on existing visualisation packages to implement forest plots [*ggforestplot*, @orchaRd; @ggforestplot], funnel plots [*metafor*, @metafor], and specification curve analysis plots [*specr*, @Masur2020] to examine the influence of analysis specifications on variation in analysis outcomes. Select functions from the above packages have been thoughtfully abstracted and coded into wrapper functions according to a principled and structured workflow (described below) for ease of application and generalisation to alternative datasets.

## "Same Data, Different Analysts: Variation in Effect Sizes Due to Analytical Decisions in Ecology and Evolutionary Biology"

The *ManyEcoEvo* dataset contains anonymised analyst responses for two research questions: one examining blue tit nestling growth (evolutionary biology) and another examining eucalyptus seedling recruitment (conservation ecology).

```{r}
#| eval: false
library(ManyEcoEvo)
library(tidyverse)
library(metafor)
```

```{r}
#| label: load-full-dataset
# Load the built-in dataset
data("ManyEcoEvo")
# Examine the basic structure
ManyEcoEvo
glimpse(ManyEcoEvo)
```

The data is structured with list-columns:

-   `data`: Contains the effect size estimates and analysis metadata for each dataset
-   `diversity_data`: Contains information about variables used by each analysis team

We will work with a subset of the blue tit dataset for illustration purposes.

```{r}
#| label: create-MWE-data
# Extract only necessary columns for illustration
key_cols <- c(
  "id_col",
  "TeamIdentifier",
  "beta_estimate",
  "beta_SE",
  "sample_size",
  "adjusted_df",
  "dataset",
  "mixed_model"
)

analysis_data <- ManyEcoEvo %>%
  mutate(
    data = map(
      data,
      ~ .x %>%
        filter(if_all(all_of(key_cols), ~ !is.na(.))) %>%
        slice_sample(n = 25) %>%
        arrange(TeamIdentifier, analysis_id)
    ),
    diversity_data = map2(
      diversity_data,
      data,
      ~ semi_join(.x, .y, by = "id_col")
    )
  )

effect_sizes <- analysis_data %>%
  pluck("data", 1) %>%
  select(all_of(key_cols))

head(effect_sizes)
```

Each row represents a unique analysis (`id_col`), potentially with multiple analyses per team (`TeamIdentifier`).

### Analysing the Blue Tit Dataset: Data preparation for meta-analysis

The first step in any meta-analysis is ensuring that the estimands of interest – in this case, effect sizes as correlation coefficients – are on the same response scale to ensure that we are not 'comparing apples to oranges.'

The meta-analyses in the ManyEcoEvo study are performed on Fishers' Z transformed correlation coefficients $Z_r$ with the function `est_to_zr()`, which is called by `standardise_response()`, adding columns of standardised estimates `Zr` and their variance `VZr`:

```{r}
# label: glimpse-effect-sizes
standardised_effect_sizes <- effect_sizes %>%
  standardise_response(estimate_type = "Zr")
standardised_effect_sizes %>%
  glimpse()
```

Now that we have standardised the analysts' estimates, we can progress to conducting our meta-analysis.

### Analysing the Blue Tit Dataset: Meta-analysis and univariate analyses

We fit a metaregression to the standardised effect sizes, using the `metafor` package, with nested random effects:

```{r}
#| label: glimpse-MA-model
standardised_effect_sizes <- standardised_effect_sizes %>%
  mutate(study_id = id_col)
ma_model <- fit_MA_mv(
  effects_analysis = standardised_effect_sizes,
  outcome_colname = "Zr",
  outcome_SE_colname = "VZr",
  estimate_type = "Zr"
)
summary(ma_model)
```

#### Calculate Inputs for Univariate Analyses

Calculate deviation of $Z_r$ from the meta-analytic mean for each analysis `id_col`, then Box-Cox transform each deviation score:

```{r}
#| label: BoxCox-deviation-scores
effect_size_data <- standardised_effect_sizes %>%
  calculate_deviation_score(
    meta_analytic_model = ma_model,
    outcome_colname = "Zr"
  ) %>%
  box_cox_transform(dataset = "blue tit", outcome_SE_colname = "VZr")
```

Calculate mean Sorensen's similarity index for each analysis:

```{r}
#| label: extract-diversity-data
# Extract diversity data
diversity_data <- analysis_data %>%
  pluck("diversity_data", 1)
mean_sorensen_data <-
  diversity_data %>%
  calculate_sorensen_diversity_index(.id = "id_col")
mean_sorensen_data
```

#### Regression on Distinctiveness of Analysis Variables

Join required `mean_diversity_indices` into `effect_size_data` and fit a univariate `glm` model of deviation scores on mean Sorensen's diversity index. Note that the returned model is fitted using *parsnip* [@parsnip2024].

```{r}
#| label: glimpse-diversity-model
effect_size_data <- effect_size_data %>%
  left_join(mean_sorensen_data, by = join_by(id_col))

diversity_model <- fit_sorensen_glm(effect_size_data)
diversity_model # View model summary
```

#### Visualise Results

Create a forest plot for the blue tit meta-analysis:

```{r}
#| label: fig-ManyEcoEvoR-forest
#| fig-cap: |
#|    Forest plots of meta-analytic estimated standardised effect sizes ($Z_r$, blue circles) and their 95% confidence intervals for each effect size included in the meta-analysis model. Results are shown for a subset of the complete blue tit dataset from @Gould2025. Meta-analytic mean effect size is shown as diamond.
#| fig-scap: "Forest plots of meta-analytic estimated standardised effect sizes for a subset of the blue tit dataset."
get_forest_plot_data(ma_model) %>%
  plot_forest(intercept = TRUE)
```

Create a marginal effects plot for the univariate analysis on the distinctiveness of variables, back-transforming the response variables for easier interpretation:

```{r}
#| label: fig-ManyEcoEvoR-Sorensen
#| fig-cap: |
#|   Fitted model of the Box-Cox transformed deviation score (deviation in the effect size from the meta-analytic mean) as a function of the mean Sorensen's index showing distinctiveness of the set of predictor variables, for a subset of the blue tit dataset from @Gould2025. Grey ribbons indicate predicted values are 95% CI's. Deviation scores have been back-transformed to the original scale.
#| fig-scap: "Scatterplot of Box-Cox transformed deviation scores as a function of the the mean Sorensen's index."
plot_effects_diversity(
  mod = diversity_model,
  df = effect_size_data,
  back_transform = TRUE
)
```

### Scaling Up

We conducted further univariate regressions to quantify the effects of different analysis features on heterogeneity among analysts' results in addition to the distinctiveness of variables in each analysis, as we illustrated above. We also performed the full analysis on a second dataset, the *Eucalyptus* dataset, and repeated the analysis on different outcome variables (point predictions from analysts' models, not shown in this manuscript), and different subsets of data (e.g., with outliers removed, with poorly rated analyses, among others).

To make conducting multiple analyses easier and to facilitate scaling up the analysis pipeline for repeated application over different analysis questions, data subsets, and outcome variables, we provide a selection of 'wrapper' functions. The wrapper frameworks utilise the *tidyverse*'s [@tidyverse2019] list-column data structures and iteration with *purrr* [@purrr2025]. We illustrate this functionality in detail within [*ManyEcoEvo*'s vignettes](https://egouldo.github.io/ManyEcoEvo/).

## Conclusion

We present *ManyEcoEvo*, an R package for reproducing, re-analysing or replicating the "Same Data, Different Analysts" study in ecology and evolution [@Gould2025]. The process of curating, analysing and visualising many-analyst style data, like that of the *ManyEcoEvo* project, is a painstaking task that requires careful computational organisation to ensure that the study results are transparent, replicable and computationally reproducible. The *ManyEcoEvo* package provides the tools to make this task more approachable, and its underlying foundations are scalable for re-analysis over different outcome variables, data subsets, or even different many-analyst datasets. Future work will incorporate other meta-analysis methods.

# Session Information

```{r}
#| label: session-info

sessioninfo::session_info()
```


# References
