---
title: "Scaling Up: working with multiple subsets or multiple datasets"
output: 
  bookdown::html_document2:
      base_format: rmarkdown::html_vignette
      fig_caption: yes
      toc: true
pkgdown:
  as_is: true
vignette: >
  %\VignetteIndexEntry{multiple_datasets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ManyEcoEvo)
```

# A tidy approach with list-columns and nested dataframes
list-columns tidy modelling approach.

## Working with Multiple datasets

Multiple different datasets that you might want to replicate the same analyses and compare. E.g. Blue tit vs. Eucalyptus. 

Demonstrate the approach of the analysis pipeline for ManyEcoEvo just using the blue tit and Eucalyptus data -> `make_viz()` etc. 

## Creating data subsets based on various exclusion principles

Generate different subsets:
- `generate_exclusion_subsets()`
- `generate_expertise_subsets()` 
- `generate_outlier_subsets()` 
- `generate_rating_subsets()` 
- `generate_yi_subsets()`
