---
title: "ManyEcoEvo: an R package for Analysing Many-analyst-style Data"
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
cache: true
tags:
  - R
  - meta-analysis
  - metaresearch
  - ecology
  - evolutionary biology
  - many-analysts
authors:
  - name: Elliot Gould
    orcid: 0000-0003-0872-7098
    email: "elliot.gould@unimelb.edu.au"
    affiliations:
      - ref: aff1
  - name: Hannah S. Fraser
    orcid: 0000-0000-0000-0000
    affiliations:
      - ref: aff1
  - name: Shinichi Nakagawa
    orcid: 0000-0000-0000-0000
    affiliations:
      - ref: aff2
  - name: Timothy H. Parker
    orcid: 0000-0000-0000-0000
    affiliations:
      - ref: aff3
citation_author: Gould and Fraser et al.
affiliations:
  - id: aff1
    name: The University of Melbourne
  - id: aff2
    name: The University of New South Wales
  - id: aff3
    name: Whitman College
bibliography: "vignettes.bib"
---

```{r}
#| echo: false
#| message: false
#| warning: false
#| cache: false
#| label: setup
library(ManyEcoEvo)
library(tidyverse)
library(metafor)
set.seed(1)
knitr::opts_chunk$set(
  # collapse = TRUE,
  comment = "#>",
  # fig.width = 7,
  # fig.height = 5,
  message = FALSE,
  warning = FALSE,
  include = TRUE,
  cache = TRUE,
  echo = TRUE,
  tidy = "styler"
)
```

## Summary

Many-analyst studies are an emerging format of crowdsourced scientific analysis [@Schweinsberg2021], in which multiple analysts independently answer the same research question using the same dataset [@Voracek2019; @Silberzahn2018]. Several high-profile studies demonstrate that results and conclusions can be highly contingent on subjective but defensible analytical decisions [@Silberzahn2018; @Botvinik-Nezer2020; @Breznau2021; @Huntington-Klein2021; @Hoogeveen2022; @Coretta2023], arising from disagreement about modelling strategy or variation in data preparation [@Voracek2019; @Simonsohn2020]. Similarly to meta-analysis, many-analyst studies treat analytical variation as the object of study, and produce rich metadata for each analysis — including model type, transformations, and variable selection — that can be used to characterise analytical uncertainty. Unlike meta-analysis, the source of variation under study is analytical rather than arising from sampling or measurement error between independent datasets.

`ManyEcoEvo` is an R package that provides a structured, reproducible workflow for conducting the statistical analyses characteristic of many-analyst studies. The package implements a pipeline that: standardises heterogeneous analyst estimates onto a common scale; fits multilevel meta-analytic models across multiple analyst submissions; computes each analyst's deviation from the meta-analytic mean; and fits a series of univariate regression models examining how analytical choices relate to deviation from the consensus estimate among teams. The pipeline is built around a nested tibble structure that allows the full sequence of operations to be mapped systematically over multiple datasets and over systematically varied analytical subsets of those datasets — for example, subsets defined by peer-review quality ratings, variable collinearity, or analyst expertise. This architecture makes repeated application of the analysis pipeline across many dataset variants a tractable problem.

Although `ManyEcoEvo` was developed for the 'Same Data, Different Analysts' study in ecology and evolutionary biology [@Gould2025], its workflow generalises to computationally analogous problems. Collaborative big-team open science projects (e.g. '[ManyBirds](http://themanybirds.com/)' @ManyBirdsProject2025; '[ManyBabies](https://manybabies.org/)' @ManyBabies; @McMillan2026, '[ManyPrimates](https://manyprimates.github.io/)' @ManyPrimates) and coordinated distributed experiments (e.g. '[NutNet](https://nutnet.org/)'; @Borer2014, '[DragNet](https://dragnetglobal.org/)' @Borer2025) produce multiple datasets that pose similar data-manipulation and analysis challenges. The package also generalises to sensitivity analyses in conventional meta-analyses, where the primary analysis is repeated across varied inclusion criteria or data subsets.

In addition to the workflow infrastructure, `ManyEcoEvo` ships the curated, analysis-ready datasets from @Gould2025 as package data objects. These include analyst effect-size estimates and metadata, out-of-sample predictions as an alternative outcome variable, peer-review ratings, and variable-use data for two research questions — one from evolutionary biology (blue tit nestling growth) and one from conservation ecology (*Eucalyptus* seedling recruitment). The packaged data enables straightforward reproduction and re-analysis of the published results.

## Statement of Need

Conducting a many-analyst study requires solving a class of data-management and analytical problems that existing meta-analytic software does not address. The primary challenge is heterogeneity: analyst submissions span a wide variety of modelling approaches, response variable transformations, and link functions, all of which must be identified, back-transformed, and standardised onto a common scale before any synthesis is possible. In a study involving over a hundred analyst teams, performing this step manually — tracing each analyst's reported transformation, applying the appropriate back-transformation, and propagating the result through to a standardised effect size or out-of-sample prediction — is extremely time-consuming and error-prone.

A second challenge is systematic repetition. Many-analyst studies - especially like @Gould2025 where different analysis questions were posed - require the full many-analyst pipeline to be repeated across multiple analytical subsets. For example, on subsets; with analyses that were rated as low quality by peer reviewers, or that contained highly collinear predictors; with extreme outliers removed; that were restricted to expert analysts. Without dedicated infrastructure, this repetition is implemented as bespoke, study-specific scripts: copies of the primary analysis, manually adapted for each subset. This approach makes it difficult to reason about what differs between subsets and what does not, obscures whether observed differences in results across subsets are attributable to the subset definition or to incidental differences in the code, and makes it hard to verify that the same analytical protocol has been applied consistently. It also makes the analysis difficult to reproduce, audit, or adapt to new data.

The target audience for `ManyEcoEvo` is therefore researchers conducting or re-analysing many-analyst studies, as well as researchers whose work involves the systematic repetition of a meta-analytic pipeline across multiple datasets or analytically defined data subsets. This includes teams running collaborative big-team open science projects where data are collected across multiple experiments, locations or analyst groups. A secondary audience is conventional meta-analysts for whom sensitivity analyses, which repeat the primary meta-analysis under varied inclusion criteria, are a routine but currently ad hoc component of their workflow.

`ManyEcoEvo` addresses these challenges by providing a reusable, programmatic workflow that treats outcome-variable harmonisation and standardisation, subset generation, and pipeline repetition as first-class operations. Standardisation and back-transformation of heterogeneous analyst estimates are handled by a family of functions that identify and apply the appropriate transformation for each analysis, making the process explicit, auditable, and reproducible. Subset generation is implemented as row-expansion of a nested tibble, so that the full downstream pipeline — meta-analysis, computation and transformation of deviation scores, and univariate regressions — propagates automatically through all dataset and subset combinations without modification to the pipeline itself. This architecture makes it straightforward to verify that the same analytical protocol has been applied consistently across all subsets, and to isolate the effect of each subset definition on results. Together, these properties support FAIR [@Wilkinson2016] and reproducible research practice in a study format that is otherwise difficult to implement in a principled way.

## State of the Field

The R ecosystem for meta-analysis is mature and well-characterised. Lortie & Filazzola [-@lortie2020] provide a structured contrast of the two most widely used packages in ecology and evolutionary biology: meta [@meta], a direct and intuitive choice for rapid implementation of general meta-analytic statistics, and `metafor` [@metafor], a more comprehensive package suited to complex multilevel models. Both provide heterogeneity estimation, visualisation tools, and publication bias diagnostics, and both are widely used across ecology, evolution, and the social sciences. For visualisation, `orchaRd` [@orchaRd] extends `metafor` with orchard plots suited to ecological effect-size data, and `metaviz` [@metaviz] provides additional forest and funnel plot variants. These packages form the analytical foundation on which `ManyEcoEvo` builds: the multilevel meta-regression at the core of the package is implemented via `metafor::rma.mv()`, and forest plots are produced using conventions established by `orchaRd` and `metaviz`. However, all of these packages share a common design assumption — that the unit of analysis is a single meta-analytic dataset, and that the researcher's task is to fit, interrogate, and visualise one model. None provide infrastructure for the upstream data-wrangling challenges specific to many-analyst data, nor for the systematic generation and propagation of analytical subsets through a complete pipeline. Contributing the `ManyEcoEvo` workflow to any of these packages would require introducing a fundamentally different data model and pipeline architecture that sits outside their intended scope.

A second category of potentially relevant tools comes from multiverse analysis and specification curve analysis. The `multiverse` package [@Sarma2023] provides an embedded domain-specific language for declaring alternative analysis paths within R and R Notebooks, implementing the multiverse analysis philosophy of @Steegen2016. `specr` [@Masur2020] provides utilities for conducting and visualising specification curve analyses [@Simonsohn2020], allowing users to set up, run, evaluate, and plot the full space of analytical specifications for a single dataset. `Boba` [@Liu2021] extends this further with an integrated domain-specific language and interactive visualisation system for authoring and interpreting multiverse analyses across multiple programming languages. These tools share with `ManyEcoEvo` the goal of characterising how analytical variation affects results, but they operate in a fundamentally different setting: they generate variation in analysis outcomes programmatically from a single analyst's declared decision space, rather than synthesising analytic variation across many analysts or teams. Consequently, they do not address the resolution of incompatibility between analyst estimates arising from different model specifications, transformations, and link functions prior to meta-analytic synthesis, nor the nested data structures that arise when analyst submissions must be aggregated and compared. The conceptual framing is also distinct: multiverse and specification curve tools ask "what would have happened if this analyst had made different choices?", whereas `ManyEcoEvo` asks "given that many analysts made different choices, what can we infer about the distribution and drivers of those differences?" Contributing the `ManyEcoEvo` pipeline to any of these tools would require introducing a meta-analytic synthesis layer and a model for resolving heterogeneous-input data that are outside their design scope entirely.

The closest precedents for the `ManyEcoEvo` approach are the study-specific analysis repositories published alongside individual many-analyst studies, such as those accompanying the 'ManyBabies' [@ManyBabies] and 'ManyPrimates' [@ManyPrimates] projects. These repositories demonstrate that the analytical problems `ManyEcoEvo` addresses recur across disciplines. However, their code is written for a single study and dataset, with no abstraction or reuse intended. Contributing to these repositories would produce more study-specific code rather than a reusable, generalisable tool. `ManyEcoEvo` was therefore developed as a standalone package because no existing tool addresses the combination of resolution of heterogeneity in outcome variables, nested data management, systematic subset generation, and reproducible pipeline repetition that many-analyst studies require.

## Research Impact

`ManyEcoEvo` was developed to support the analysis reported in @Gould2025, a large-scale many-analyst study in ecology and evolutionary biology in which 174 analyst teams independently investigated the same two research questions. That paper has accumulated 54 citations, over 40,000 accesses, and an Altmetric score of 347 since publication, indicating substantial engagement from the research community. The packaged datasets and workflow have been independently used by @Knief2025, who conducted a targeted multiverse re-analysis of the blue tit dataset from @Gould2025, demonstrating that the curated data objects shipped with `ManyEcoEvo` are sufficient to support independent re-analysis and extension of the published results.

The package is archived on Zenodo with a citable DOI, and has accumulated 670 views and 150 downloads across all versions. The GitHub repository (<https://github.com/egouldo/ManyEcoEvo>) has received 264 clones from 91 unique cloners in the 14 days prior to submission, suggesting active and ongoing interest from the research community. Package documentation is maintained at a dedicated `pkgdown` site (@Wickham2024, <https://egouldo.github.io/ManyEcoEvo/>), and the repository follows community standards for R package development [@wickham2023r], including a ticketed issue tracker, pull request workflow, and [contributor guidelines](#0 "Contributor Guidelines: https://github.com/egouldo/ManyEcoEvo?tab=contributing-ov-file#") including a [code of conduct](#0 "Code of conduct: https://github.com/egouldo/ManyEcoEvo/blob/main/CODE_OF_CONDUCT.md"). These infrastructure collectively indicate that the package is ready for community use and external contribution.

## Software Design

`ManyEcoEvo` is built around a nested tibble as its central data structure, in which each row represents a dataset or analytical subset, and the analyst-level data for that row is stored in list-columns rather than as flat rows. This design was chosen deliberately because it maps naturally onto the problem structure of many-analyst studies: the same sequence of operations must be applied repeatedly and consistently across multiple datasets and multiple variants of those datasets. Many-analyst data also has an inherently hierarchical structure — a single research question may receive multiple analyses from the same team, and each analysis may be reviewed by multiple reviewers — that requires grouped operations to handle correctly. By representing each dataset or subset as a row with analyst-level data stored in a list-column, grouping and iteration are expressed cleanly and explicitly: the full pipeline can be applied as row-wise transformations using `purrr::map()` variants from the tidyverse [@tidyverse2019], and grouped operations within a dataset are handled naturally by the nested structure without requiring fragile index-based bookkeeping. This makes the pipeline both concise and legible: the structure of the code mirrors the structure of the analysis, and adding a new dataset or subset requires only an additional row rather than a modification to the pipeline itself.

A flat data frame approach was considered and rejected. While flat data frames are straightforward for simple operations, they become unwieldy when joining heterogeneous nested objects — such as data on the distinctiveness of analysis variables, peer-review ratings, and back-transformed point estimates — across datasets for downstream plotting and analysis. They are also poorly suited to the programmatic generation of multiple subset variants, where each variant requires its own correctly filtered copy of several related data objects. Managing these relationships in a flat structure requires either duplicating data or maintaining fragile index-based joins, both of which are error-prone at the scale of a many-analyst study. The nested tibble approach keeps related data objects co-located in the same row, making the relationships between them explicit and reducing the risk of misalignment.

The pipeline also makes a deliberate distinction between modular analysis-level functions, such as `est_to_zr()`, `calculate_deviation_score()`, and `box_cox_transform()`, and wrapper functions like `meta_analyse_datasets()`, which coordinate the full sequence for a given dataset. This two-level architecture means that users who wish to apply only part of the pipeline, substitute their own standardisation method, or adapt the workflow to a different study design can do so by calling the modular functions directly, without being constrained by the wrapper. The wrapper functions exist for convenience and reproducibility, not as the only entry point.

Back-transformation of heterogeneous analyst point estimates is handled via simulation: for each analyst, a distribution of estimates is generated by sampling from a normal distribution parameterised by the reported point estimate and standard error, the appropriate inverse transformation is applied to the full distribution, and the mean and standard deviation of the transformed distribution are returned. This approach correctly propagates uncertainty through non-linear transformations, at the cost of some computational overhead relative to analytical delta-method approximations. The simulation approach was preferred because it is general — it handles any transformation for which an inverse exists — and because it makes the back-transformation procedure transparent and independently verifiable.

The relationship between the software and the associated publication warrants a brief note on versioning. Version 2.7.6 of `ManyEcoEvo`, archived on Zenodo with a versioned DOI and tagged on GitHub, corresponds exactly to the software used to produce the results reported in @Gould2025. The current development version (2.8.0.9000) extends this release and is distinguished from it by both the Zenodo version DOI and the GitHub release tag. Users wishing to exactly reproduce the results of @Gould2025 should install the archived version; users wishing to apply the workflow to new data should use the current release.

Package dependencies are managed with `renv` [@renv2024], which records the full set of package versions in a project-specific library, ensuring that the computational environment can be restored exactly on a clean install. The analysis pipeline is implemented using the `targets` package [@landau2021], which makes the dependency structure between pipeline steps explicit and provides caching so that only outdated steps are re-executed when inputs change. Data integrity throughout the pipeline is enforced using `pointblank` [@pointblank2025], which performs programmatic validation checks on data structures at key steps, catching malformed inputs early and making data quality expectations explicit and auditable. A formal unit test suite is implemented using `testthat` [@testthat2011]. Full instructions for a clean install and pipeline execution are provided in the package README at <https://github.com/egouldo/ManyEcoEvo>.

## Software Demonstration

`ManyEcoEvo` is distributed with two curated, analysis-ready datasets from @Gould2025 as package data objects: `ManyEcoEvo`, containing analyst effect-size estimates, and `ManyEcoEvo_yi`, containing out-of-sample predictions. Both use a nested tibble structure in which each row corresponds to a dataset and analyst-level data is stored in list-columns:

```{r}
#| label: load-data
library(ManyEcoEvo)
library(tidyverse)
library(metafor)

data("ManyEcoEvo")
ManyEcoEvo
```

Each element of the list-column `data` is a tibble of analyst submissions for that dataset; `diversity_data` stores the predictor-variable data used in analysis of the influence of distinctiveness of model specification on variation in effect-sizes. This structure means all downstream operations are expressed as row-wise transformations using `purrr::map()` variants, and the pipeline never needs to unpack and repack a flat data frame.

Back-transformation and standardisation of heterogeneous analyst estimates onto a common scale is handled by `standardise_response()`, which identifies the appropriate transformation for each analysis and returns standardised effect sizes and their variances:

```{r}
#| label: standardise
blue_tit_data <- ManyEcoEvo %>%
  filter(dataset == "blue tit") %>%
  pluck("data", 1)

standardised <- blue_tit_data %>%
  standardise_response(estimate_type = "Zr")

standardised %>%
  select(id_col, Zr, VZr) %>%
  head()
```

The full analysis pipeline — meta-analytic model fitting, deviation score computation and transformation, and univariate regressions — is coordinated by the wrapper function `meta_analyse_datasets()`, which executes the complete sequence for each row of the nested tibble and returns results as additional list-columns. Because the function iterates over rows, passing a two-row object produces two complete sets of results with no change to the pipeline call:

```{r}
#| label: full-pipeline
#| message: false
#| warning: false
results <-
  ManyEcoEvo %>%
  prepare_response_variables(
    estimate_type = "Zr",
    dataset_standardise = c("blue tit", "eucalyptus")
  ) %>%
  compute_MA_inputs(estimate_type = "Zr") %>%
  meta_analyse_datasets(
    outcome_variable = "Zr",
    outcome_SE = "VZr"
  )

results %>%
  select(dataset, MA_mod, sorensen_glm, box_cox_rating_cont)
```

Analyses on subsets of the complete analyst datasets are implemented by expanding the nested tibble with additional rows, one per subset, using the `generate_*()` family of functions. The full meta-analysis pipeline then propagates through all combinations without modification:

```{r}
#| label: subsets
#| message: false
#| warning: false
results_with_subsets <-
  ManyEcoEvo %>%
  prepare_response_variables(
    estimate_type = "Zr",
    dataset_standardise = c("blue tit", "eucalyptus")
  ) %>%
  generate_exclusion_subsets(estimate_type = "Zr") %>%
  generate_rating_subsets() %>%
  compute_MA_inputs(estimate_type = "Zr") %>%
  meta_analyse_datasets(
    outcome_variable = "Zr",
    outcome_SE = "VZr"
  )

results_with_subsets %>%
  distinct(dataset, exclusion_set, publishable_subset)
```

A complete worked example, including forest plots, marginal effects plots, and extraction of model fit statistics, is provided in the package vignette at <https://egouldo.github.io/ManyEcoEvo/>.

# References
