---
title: "ManyEcoEvo: an R package for Analysing Many-analyst-style Data"
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
tags:
  - R
  - meta-analysis
  - metaresearch
  - ecology
  - evolutionary biology
  - many-analysts
authors:
  - name: Elliot Gould
    orcid: 0000-0003-0872-7098
    email: "elliot.gould@unimelb.edu.au"
    affiliations:
      - ref: aff1
  - name: Hannah S. Fraser
    orcid: 0000-0000-0000-0000
    affiliations:
      - ref: aff1
  - name: Shinichi Nakagawa
    orcid: 0000-0000-0000-0000
    affiliations:
      - ref: aff2
  - name: Timothy H. Parker
    orcid: 0000-0000-0000-0000
    affiliations:
      - ref: aff3
citation_author: Gould and Fraser et al.
affiliations:
  - id: aff1
    name: The University of Melbourne
  - id: aff2
    name: The University of New South Wales
  - id: aff3
    name: Whitman College
bibliography: "vignettes.bib"
---

```{r}
#| echo: false
#| message: false
#| warning: false
#| cache: false
#| label: setup
library(ManyEcoEvo)
library(tidyverse)
library(metafor)
set.seed(1)
knitr::opts_chunk$set(
  # collapse = TRUE,
  comment = "#>",
  # fig.width = 7,
  # fig.height = 5,
  message = FALSE,
  warning = FALSE,
  include = TRUE,
  cache = TRUE,
  echo = TRUE,
  tidy = "styler"
)

```


## Summary

Many-analyst studies are an emerging format of crowd-sourced scientific analysis [@Schweinsberg2021], where analysts simultaneously, but independently, answer the same research question, using the same dataset [@Voracek2019; @Silberzahn2018]. Several high-profile many-analyst studies demonstrate how results and conclusions may be highly contingent on subjective, and defensible but arbitrary analysis decisions [@Silberzahn2018; @Botvinik-Nezer2020; @Breznau2021]. Analytic flexibility may be due to genuine disagreement among researchers or analytic uncertainty in choices such as which observations to include and what data-analysis strategy is most appropriate to answer a given research question with the data at hand [@Voracek2019; @Simonsohn2020, figure 1].

Similarly to meta-analysis, many-analyst studies treat analytical variation as the object of study. Yet the two study formats differ in the source of variation they examine, their study goals with respect to analysing variation, assumptions about analytical choices.

Meta-analyses focus on synthesising data from multiple independent studies to calculate combined effect sizes, while many-analyst studies aim to characterise and understand analytic uncertainty, applying meta-analytic methods to distributions of effect sizes across individual or teams of independent researchers. Consequently, meta-analyses over individual studies assume the source of variation is due to different samples or populations, while many-analyst studies consider analytic variation as an additional source of heterogeneity in results. Many-analyst studies produce rich metadata for each analysis, including model type, transformations, variable selection or other analytic choices, that may either be used within meta-analyses or else as predictors or covariates within additional analyses that aim to characterise analytic uncertainty. Further, the two study types may differ in the scale and repetition involved; many-analyst studies may systematically repeat the same analyses over multiple datasest, multiple analytical subsets, or multiple outcome variables - for meta-analyses, such analyses are usually conducted as sensitivity analyses and are performed in an ad hoc, and poorly documented manner. For example, a conventional meta-analysis ay ask, "what happens to the pooled estimate if we remove outliers / restrict to high-quality studies / exclude a taxonomic group?"

### Statement of Need

Existing meta-analytic software within the R ecosystem are designed around fitting and interrogating a single model on a single dataset, with no infrastructure for systematic, programatic repetition across multiple datasets or subsets of analyses. While packages such as `metafor`, `meta`, and `orchaRd` provide excellent tools for conventional meta-analysis, none are designed to handle the specific data structure and analytical questions that arise in many-analysts style research. Many-analyst studies might share study-specific code, but it is not designed for application to new data (e.g. [github.com/manybabies/mb1-analysis-public](https://github.com/manybabies/mb1-analysis-public), [github.com/many-speech-analyses/many_analyses](https://github.com/many-speech-analyses/many_analyses)). Additionally, dedicated software designed to analyse and visualise analytic variation, for example decision analytic graphs or specification curves, are similarly built for application to a single dataset. Consequently, we developed ManyEcoEvo to facilitate the analysis of the 'Same Data, Many Analysts' study in ecology and evolutionary biology [@Gould2025], in which 174 analyst teams, comprising 246 analysts, independently investigated the answers to two prespecified research questions and datasets, one from conservation ecology, and one from evolutionary biology.

ManyEcoEvo provides a reproducible data pipeline and a multi-level wrapper framework for running meta-analyses programmatically across multiple datasets and over systematically varied analytical subsets of those datasets. The package provides a reusable workflow infrastructure that also generalises to sensitivity analyses in conventional meta-analyses. The ManyEcoEvo package occupies a distinctive niche in the R ecosystem that is not addressed by existing meta-analysis tooling. Its secondary contribution is to provide a curated, analysis-ready dataset to facilitate reproductions or re-analyses of @Gould2025.

## Dataset Description


The *ManyEcoEvo* dataset contains analyst responses for two different research questions, each with their own analysis dataset: one examining blue tit nestling growth (evolutionary biology) and another examining eucalyptus seedling recruitment (conservation ecology).

```{r}
#| eval: false
library(ManyEcoEvo)
library(tidyverse)
library(metafor)
```

```{r}
#| label: load-full-dataset
# Load the built-in dataset
data("ManyEcoEvo")
# Examine the basic structure
ManyEcoEvo
glimpse(ManyEcoEvo)
```

The data is structured with list-columns:

-   `data`: Contains the effect size estimates and analysis metadata for each dataset
-   `diversity_data`: Contains information about variables used by each analysis team

We will work with a subset of the blue tit dataset for illustration purposes.

```{r}
#| label: create-MWE-data
# Extract only necessary columns for illustration
key_cols <- c(
  "id_col",
  "TeamIdentifier",
  "beta_estimate",
  "beta_SE",
  "sample_size",
  "adjusted_df",
  "dataset",
  "mixed_model"
)

analysis_data <- ManyEcoEvo %>%
  mutate(
    data = map(
      data,
      ~ .x %>%
        filter(if_all(all_of(key_cols), ~ !is.na(.))) %>%
        slice_sample(n = 25) %>%
        arrange(TeamIdentifier, analysis_id)
    ),
    diversity_data = map2(
      diversity_data,
      data,
      ~ semi_join(.x, .y, by = "id_col")
    )
  )

effect_sizes <- analysis_data %>%
  pluck("data", 1) %>%
  select(all_of(key_cols))

head(effect_sizes)
```

Each row represents a unique analysis (`id_col`), potentially with multiple analyses per team (`TeamIdentifier`).

## Calculating meta-analytic results from raw estimates with the Blue tit dataset

Here we introduce analysis-level functions in ManyEcoEvo demonstrating on a single dataset a typical analysis pipeline which moves from data cleaning and harmonisation, meta-analysis and analysis of sources of heterogeneity, and visualisation and reporting of results. Each of these functions is modular and usable independently of our package, so users can substitute their own standardisation method, their own model-fitting function, or their own deviation score approach.

### Data preparation for meta-analysis

The first step in any meta-analysis is ensuring that the estimands of interest – in this case, effect sizes as correlation coefficients – are on the same response scale to ensure that we are not 'comparing apples to oranges.'

The meta-analyses in the ManyEcoEvo study are performed on Fishers' Z transformed correlation coefficients $Z_r$ with the function `est_to_zr()`, which is called by `standardise_response()`, adding columns of standardised estimates `Zr` and their variance `VZr`:

```{r}
# label: glimpse-effect-sizes
standardised_effect_sizes <- effect_sizes %>%
  standardise_response(estimate_type = "Zr")
standardised_effect_sizes %>%
  glimpse()
```

Now that we have standardised the analysts' estimates, we can progress to conducting our meta-analysis.

### Meta-analysis and univariate analyses

We fit a metaregression to the standardised effect sizes, using the `metafor` package, with nested random effects:

```{r}
#| label: glimpse-MA-model
standardised_effect_sizes <- standardised_effect_sizes %>%
  mutate(study_id = id_col)
ma_model <- fit_MA_mv(
  effects_analysis = standardised_effect_sizes,
  outcome_colname = "Zr",
  outcome_SE_colname = "VZr",
  estimate_type = "Zr"
)
summary(ma_model)
```

#### Calculate Inputs for Univariate Analyses

Calculate deviation of $Z_r$ from the meta-analytic mean for each analysis `id_col`, then Box-Cox transform each deviation score:

```{r}
#| label: BoxCox-deviation-scores
effect_size_data <- standardised_effect_sizes %>%
  calculate_deviation_score(
    meta_analytic_model = ma_model,
    outcome_colname = "Zr"
  ) %>%
  box_cox_transform(dataset = "blue tit", outcome_SE_colname = "VZr")
```

Calculate mean Sorensen's similarity index for each analysis:

```{r}
#| label: extract-diversity-data
# Extract diversity data
diversity_data <- analysis_data %>%
  pluck("diversity_data", 1)
mean_sorensen_data <-
  diversity_data %>%
  calculate_sorensen_diversity_index(.id = "id_col")
mean_sorensen_data
```

#### Regression on Distinctiveness of Analysis Variables

Join required `mean_diversity_indices` into `effect_size_data` and fit a univariate `glm` model of deviation scores on mean Sorensen's diversity index. Note that the returned model is fitted using *parsnip* [@parsnip2024].

```{r}
#| label: glimpse-diversity-model
effect_size_data <- effect_size_data %>%
  left_join(mean_sorensen_data, by = join_by(id_col))

diversity_model <- fit_sorensen_glm(effect_size_data)
diversity_model # View model summary
```

#### Visualise Results

Create a forest plot for the blue tit meta-analysis:

```{r}
#| label: fig-ManyEcoEvoR-forest
#| fig-cap: |
#|    Forest plots of meta-analytic estimated standardised effect sizes ($Z_r$, blue circles) and their 95% confidence intervals for each effect size included in the meta-analysis model. Results are shown for a subset of the complete blue tit dataset from @Gould2025. Meta-analytic mean effect size is shown as diamond.
#| fig-scap: "Forest plots of meta-analytic estimated standardised effect sizes for a subset of the blue tit dataset."
get_forest_plot_data(ma_model) %>%
  plot_forest(intercept = TRUE)
```

Create a marginal effects plot for the univariate analysis on the distinctiveness of variables, back-transforming the response variables for easier interpretation:

```{r}
#| label: fig-ManyEcoEvoR-Sorensen
#| fig-cap: |
#|   Fitted model of the Box-Cox transformed deviation score (deviation in the effect size from the meta-analytic mean) as a function of the mean Sorensen's index showing distinctiveness of the set of predictor variables, for a subset of the blue tit dataset from @Gould2025. Grey ribbons indicate predicted values are 95% CI's. Deviation scores have been back-transformed to the original scale.
#| fig-scap: "Scatterplot of Box-Cox transformed deviation scores as a function of the the mean Sorensen's index."
plot_effects_diversity(
  mod = diversity_model,
  df = effect_size_data,
  back_transform = TRUE
)
```

## Scaling up: Systematic Analysis Across Datasets and Analytical Subsets

- Explain the nested tibble data structure as the object that the wrappers operate on — dataset, data, diversity_data as list-columns, grouped by exclusion_set, estimate_type, etc.

- Explain generate_exclusion_subsets() and related functions as expanding the nested structure by systematically generating new analytical subsets, so that the full downstream pipeline propagates through all subsets automatically. Draw the analogy to sensitivity analyses in conventional meta-analysis.



- Explain meta_analyse_datasets() as the main coordination function — fitting models, computing deviation scores, applying transformations, and fitting downstream regressions in a single call across all dataset-by-subset combinations.

- Show a minimal code example using prepare_response_variables() -> compute_MA_inputs() -> meta_analyse_datasets() -> make_viz() (pull from existing vignette).

- Note that filter_vars, outcome_variable, and VZ_colname as named lists are the user-facing interface points that make the framework generalisable to new datasets without modifying source code.

## Integration with `targets::`

The existing article does not mention targets:: at all, which is a notable omission given that the entire pipeline is implemented using targets ...

Add short section noting that the package is designed around targets:: for pipeline reproducibility

strengthens the reproducibility claim and is relevant to the JOSS audience.

tar_make() reproduces the entire analysis from raw data to results, and that the pipeline architecture makes the computational dependency structure explicit and auditable.

## Conclusion

ManyEcoEvo provides a structured workflow that maps analytical operations systematically across datasets and data subsets. The packages wrapper architecture generalises to conventional meta-analyses, where sensitivity analyses repeat the meta-analysis across inclusion criteria or data subsets. Its contribution is largely architectural, making systematic, reproducible, large-scale repetition of the same analysis pipeline across many dataset variants a tractable, first-class operation rather than an ad hoc afterthought. Finally, it is also a data package for the 'Same Data, Many Analysts' study in Ecology and Evolution [@Gould2025].
